---
title: "Reproducible Research - Course Project 1"
author: "Junaid Ishaq"
date: "1/24/2020"
output:
  html_document:
    keep_md: true
    theme: united
    highlight: default
---

&nbsp;

## Main Tasks
1. Loading and preprocessing the data
    + Load the data
    + Process/transform the data (if necessary)
2. Calculating mean total number of steps taken per day
    + Total number of steps taken per day
    + Make a histogram of the total number of steps taken each day
    + Report the mean and median of the total number of steps taken per day
3. Average daily activity pattern
    + Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
    + Determine 5-minutes interval containing the maximum number of steps
4. Imputing missing values
    + Report the total number of missing values in the dataset
    + Fill in all of the missing values
    + Create a new dataset that is equal to the original dataset but with the missing data filled in
    + Make a histogram of the total number of steps taken each day
    + Report the mean and median of the total number of steps taken per day
    + Report the difference in mean, median and total number of steps in new and old datasets
5. Differences in activity patterns between weekdays and weekends
    + Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
    + Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

&nbsp;

## 1. Loading and preprocessing the data
### 1.1 Data loading
```{r echo=TRUE}
library(knitr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(kableExtra, warn.conflicts = FALSE)

setwd("C:/Users/junai/Desktop")
activity <- read.csv("activity.csv")
```

### 1.2 Data preprocessing/transforming
Not required!

&nbsp;

## 2. Calculating mean total number of steps taken per day
### 2.1 Total number of steps
```{r echo=TRUE}
total_steps <- sum(activity$steps, na.rm = TRUE)
cat("Total number of steps: ", total_steps)
```

### Reporting the total, mean and median for steps taken per day
```{r results = "asis"}
# calculate steps total, mean, median and number of NAs for each day
date_stats <- activity %>%
              group_by(date) %>%
              summarise(Total = sum(steps, na.rm=TRUE),
                        Mean = mean(steps, na.rm=TRUE),
                        Median = median(steps, na.rm=TRUE),
                        NAs = sum(is.na(steps)))

date_stats %>%
  knitr::kable(format="html", digits=0) %>%
  kable_styling(bootstrap_options = "striped",
              full_width = F,
              position = "center")
```

&nbsp;

### 2.2 Histogram of total number of steps
```{r echo=TRUE, fig.align="center"}
g <-ggplot(date_stats, aes(x = Total))
g + geom_histogram(binwidth=2000, color="white", aes(fill=..count..)) +
    geom_rug(sides="t") +
    labs(title="Frequency of steps taken per day", x = "Steps", y = "Days") +
    scale_fill_gradient("Count", low="yellow", high="red") +
    scale_y_continuous(breaks = seq(0,12,2)) +
    scale_x_continuous(breaks = seq(0, 23000, 4000)) +
    theme(plot.title = element_text(face = "bold", color = "black", size = 11, hjust=0.5),
          axis.title = element_text(face = "bold", color = "black", size = 10),
          axis.text.x = element_text(face = "bold", color = "black", size = 10),
          axis.text.y = element_text(face = "bold", color = "black", size = 10),
          legend.title = element_blank(),
          legend.text = element_text(face = "bold", color = "black", size = 10),
          legend.position = c(0.85,0.8),
          legend.direction = "horizontal")
```

&nbsp;

### 2.3 Mean and median for total number of steps taken per day
```{r echo=TRUE}
cat("Mean for total number of steps:   ", round(mean(date_stats$Total),0))
cat("Median for total number of steps: ", median(date_stats$Total))
```

&nbsp;

## 3. Average daily activity pattern
### 3.1 Time series plot of the 5-minute interval
```{r echo=TRUE, fig.align="center"}
interval_stats <- activity %>%
                  group_by(interval) %>%
                  summarise(Total = sum(steps, na.rm=TRUE),
                            Mean = mean(steps, na.rm=TRUE),
                            Median = median(steps, na.rm=TRUE))

# Line plot
g <-ggplot(interval_stats, aes(x = interval, y = Mean))
g + geom_line(color = "blue") +
    labs(title="Average number of steps per interval", x = "Interval", y = "Steps") +
    theme(plot.title = element_text(face = "bold", color = "black", size = 11, hjust=0.5),
          axis.title = element_text(face = "bold", color = "black", size = 10),
          axis.text.x = element_text(face = "bold", color = "black", size = 10),
          axis.text.y = element_text(face = "bold", color = "black", size = 10))
```

&nbsp;

### 3.2 Determining the interval with maximum number of steps
```{r echo=TRUE}
max <- which.max(interval_stats$Mean)
cat("Interval with maximum number of steps: ", interval_stats$interval[max])
```

&nbsp;

## 4. Imputing missing values
### 4.1 Determining and reporting total number of missing values in the dataset
```{r echo=TRUE}
cat("Total number of missing values = ", sum(is.na(activity$steps)))
```

&nbsp;

### 4.2 Create a new dataset
```{r echo=TRUE}
activity_noNAs <- activity
```

&nbsp;

### 4.3 Replacing missing values in the new dataset with the average value for each interval
```{r echo=TRUE}
ind <- which(is.na(activity$steps)) # indices for NAs
activity_noNAs$steps[ind] <- interval_stats$Mean # replace NAs in the new dataset
```

&nbsp;

### 4.4 Histogram of total number of steps for the new dataset
```{r echo=TRUE, fig.align="center"}
date_stats_noNAs <- activity_noNAs %>%
                    group_by(date) %>%
                    summarise(Total = sum(steps, na.rm=TRUE),
                              Mean = mean(steps, na.rm=TRUE),
                              Median = median(steps, na.rm=TRUE))

g <-ggplot(date_stats_noNAs, aes(x = Total))
g + geom_histogram(binwidth=2000, color="white", aes(fill=..count..)) +
    geom_rug(sides="t") +
    labs(title="Frequency of steps taken per day", x = "Steps", y = "Days") +
    scale_fill_gradient("Count", low="yellow", high="red") +
    scale_y_continuous(breaks = seq(0,24,4)) +
    scale_x_continuous(breaks = seq(0, 23000, 4000)) +
    theme(plot.title = element_text(face = "bold", color = "black", size = 11, hjust=0.5),
          axis.title = element_text(face = "bold", color = "black", size = 10),
          axis.text.x = element_text(face = "bold", color = "black", size = 10),
          axis.text.y = element_text(face = "bold", color = "black", size = 10),
          legend.title = element_blank(),
          legend.text = element_text(face = "bold", color = "black", size = 10),
          legend.position = c(0.85,0.8),
          legend.direction = "horizontal")
```

&nbsp;

### 4.5 Mean and median for total number of steps in the new dataset
```{r echo=TRUE}
cat("New mean for total number of steps:   ", round(mean(date_stats_noNAs$Total),0))
cat("New median for total number of steps: ", round(median(date_stats_noNAs$Total),0))
total_steps_new <- round(sum(activity_noNAs$steps),0)
cat("Total number of steps in new dataset:  ", total_steps_new)
```

&nbsp;

### 4.6 Differences in mean, median and total number of steps between the new and old datasets
```{r echo=TRUE}
mean_diff <- mean(date_stats_noNAs$Total) - mean(date_stats$Total)
median_diff <- median(date_stats_noNAs$Total) - median(date_stats$Total)
total_diff <- total_steps_new - total_steps
```
```{r echo=TRUE}
cat("Difference in mean:   ", round(mean_diff,0))
cat("Difference in median: ", round(median_diff,0))
cat("Difference in total number of steps:   ", total_diff)
```

&nbsp;

## 5. Activity patterns for weekdays and weekends
### 5.1 Creating a new factor variable for weekdays and weekends
```{r echo=TRUE}
dates <- as.Date(activity$date)
Day <- weekdays(dates)

activity_noNAs <- cbind(activity_noNAs, Day, Day_type = Day)

wd <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
we <- c("Saturday", "Sunday")

activity_noNAs$Day_type <- as.character(activity_noNAs$Day_type)
activity_noNAs$Day_type[which(Day %in% wd)] <- "Weekday"
activity_noNAs$Day_type[which(Day %in% we)] <- "Weekend"
activity_noNAs$Day_type <- as.factor(activity_noNAs$Day_type)
```

&nbsp;

### 5.2 Plotting the average of each interval for weekdays and weekends
```{r echo=TRUE, fig.align="center"}
wday_avg <- activity_noNAs %>%
            filter(Day_type=="Weekday") %>%
            group_by(interval) %>%
            summarise(Average = mean(steps))
wday_avg$Day <- as.factor("Weekday")

wend_avg <- activity_noNAs %>%
            filter(Day_type=="Weekend") %>%
            group_by(interval) %>%
            summarise(Average = mean(steps))
wend_avg$Day <- as.factor("Weekend")

wday_wend <- rbind(wday_avg, wend_avg)

# Line plot for weekdays and weekends
g <-ggplot(data = wday_wend, aes(x = interval, y = Average))
g + geom_line(color = "blue") +
    facet_grid(Day ~ ., scales = "fixed") +
    labs(title = "Average number of steps per interval", x = "Interval", y = "Steps") +
    theme(plot.title = element_text(face = "bold", color = "black", size = 11, hjust=0.5),
          axis.title = element_text(face = "bold", color = "black", size = 10),
          axis.text.x = element_text(face = "bold", color = "black", size = 10),
          axis.text.y = element_text(face = "bold", color = "black", size = 10),
          strip.text = element_text(face = "bold", color = "black", size = 10))
```
